E1 Testing - Bayesian Mixed Models

### 2023-10-26 Manuscript/manuscript.qmd
add intro and delosh figure

---
title: "HTW"
short-title: "Variability and Extrapolation"
date: 10-11-2023
author:
- name: Thomas Gorman
  affiliation: Indiana University
  url: www.tegorman13.github.io
  email: tegorman@iu.edu
  orcid: 0000-0001-5366-5442
- name: Rob Goldstone
  affiliation: Indiana University
  url: www.pc.cogs.indiana.edu 
  email: rgoldsto@indiana.edu
  orcid: 0000-0001-8357-8358
abstract: |
  In project 1, we applied model-based techniques to quantify and control for the similarity between training and testing experience, which in turn enabled us to account for the difference between varied and constant training via an extended version of a similarity based generalization model. In project 2, we will go a step further, implementing a full process model capable of both 1) producing novel responses and 2) modeling behavior in both the learning and testing stages of the experiment. Project 2 also places a greater emphasis on extrapolation performance following training - as varied training has often been purported to be particularly beneficial in such situations. 
keywords:
  - Learning Generalization
  - Function Learning
  - Visuomotor learning
  - Training Variability
code-repo: "Access the code, data, and analysis at <https://github.com/tegorman13/htw>"
bibliography: ../Assets/Bib/HTW.bib
link-citations: true
keep-md: true
---

# Introduction

In project 1, we applied model-based techniques to quantify and control for the similarity between training and testing experience, which in turn enabled us to account for the difference between varied and constant training via an extended version of a similarity based generalization model. In project 2, we will go a step further, implementing a full process model capable of both 1) producing novel responses and 2) modeling behavior in both the learning and testing stages of the experiment. Project 2 also places a greater emphasis on extrapolation performance following training - as varied training has often been purported to be particularly beneficial in such situations. Extrapolation has long been a focus of the literature on function learning [@brehmerHypothesesRelationsScaled1974; @carrollFunctionalLearningLearning1963]. Central questions of the function learning literature have included the relative difficulties of learning various functional forms (e.g. linear vs.bilinear vs. quadratic), and the relative effectiveness of rule-based vs. association-based exemplar models vs. various hybrid models [@bottNonmonotonicExtrapolationFunction2004; @deloshExtrapolationSineQua1997; @jonesActiveFunctionLearning2018; @kalishPopulationLinearExperts2004; @mcdanielConceptualBasisFunction2005; @mcdanielPredictingTransferPerformance2009]. However the issue of training variation has received surprisingly little attention in this area.

@@

## Procedure
Upon arrival at the laboratory, participants were provided with a description of the experiment and signed informed consent forms. They were then seated in front of a computer equipped with a mouse and were given instructions on how to perform the "Hit The Wall" (HTW) visuomotor extrapolation task.

>>>>

---
title: "HTW"
short-title: "Variability and Extrapolation"
date: 10-11-2023
author:
- name: Thomas Gorman
  affiliation: Indiana University
  url: www.tegorman13.github.io
  email: tegorman@iu.edu
  orcid: 0000-0001-5366-5442
- name: Rob Goldstone
  affiliation: Indiana University
  url: www.pc.cogs.indiana.edu 
  email: rgoldsto@indiana.edu
  orcid: 0000-0001-8357-8358
abstract: |
  In project 1, we applied model-based techniques to quantify and control for the similarity between training and testing experience, which in turn enabled us to account for the difference between varied and constant training via an extended version of a similarity based generalization model. In project 2, we will go a step further, implementing a full process model capable of both 1) producing novel responses and 2) modeling behavior in both the learning and testing stages of the experiment. Project 2 also places a greater emphasis on extrapolation performance following training - as varied training has often been purported to be particularly beneficial in such situations. 
keywords:
  - Learning Generalization
  - Function Learning
  - Visuomotor learning
  - Training Variability
code-repo: "Access the code, data, and analysis at <https://github.com/tegorman13/htw>"
bibliography: ../Assets/Bib/Dissertation.bib
link-citations: true
keep-md: true
---

{{< include ../Misc/Intro.qmd >}}

@@

## Procedure

### 2023-10-26 Misc/Intro.qmd
add intro and delosh figure

In project 1, I applied model-based techniques to quantify and control for the similarity between training and testing experience, which in turn enabled us to account for the difference between varied and constant training via an extended version of a similarity based generalization model. In project 2, we will go a step further, implementing a full process model capable of both 1) producing novel responses and 2) modeling behavior in both the learning and testing stages of the experiment. Project 2 also places a greater emphasis on extrapolation performance following training - as varied training has often been purported to be particularly beneficial in such situations. Extrapolation has long been a focus of the literature on function learning [@brehmerHypothesesRelationsScaled1974; @carrollFunctionalLearningLearning1963]. Central questions of the function learning literature have included the relative difficulties of learning various functional forms (e.g. linear vs.bilinear vs. quadratic), and the relative effectiveness of rule-based vs. association-based exemplar models vs. various hybrid models [@bottNonmonotonicExtrapolationFunction2004; @deloshExtrapolationSineQua1997; @jonesActiveFunctionLearning2018; @kalishPopulationLinearExperts2004; @mcdanielConceptualBasisFunction2005; @mcdanielPredictingTransferPerformance2009]. However the issue of training variation has received surprisingly little attention in this area.

The study of human function learning investigates how people learn to map continuous input values to continuous output values based on examples. Early pioneering work by Carroll (1963) proposed that individuals abstract functional rules from examples to guide generalization to novel inputs. In his experiments, participants associated stimulus magnitudes to response magnitudes that followed linear, quadratic, or random functions. Results showed that people could extrapolate responses to novel inputs by applying the underlying rule, rather than simply matching responses to the most similar input as proposed by associative models.

Building on Carroll's ideas, Brehmer (1974) compared how well people could learn different types of functions, finding that linear functions were easier to acquire than nonlinear functions. Brehmer proposed a hypothesis testing model whereby learners begin with an expectation of a positive linear trend and adjust their functional hypothesis given sufficient evidence.

More recently, Koh and Meyer (1991) developed an adaptive regression model that incorporated a simplicity bias to account for patterns in early learning. Their model assumes learners treat stimuli and responses as logarithmically scaled and acquire polynomial rules relating the subjective stimulus and response scales.

Together, these early rule-based accounts proposed that people do not simply associate inputs to outputs but infer abstract rules to guide generalization. While subsequent work revealed limits in pure rule-based models, the pioneering work of Carroll, Brehmer, and Koh established the importance of rules and abstraction in theories of human function learning.

The study of human function learning delves into the process of learning by which individuals navigate the mapping of continuous input values to continuous output values, based on an array of examples. Foundational studies, such as those carried out by Busemeyer, Byun, DeLosh, and McDaniel in 1997, have revealed a hierarchy in function learning difficulty. Positive linear functions emerge as the most readily learned, closely followed by negative linear functions, monotonic nonlinear functions, and then non-monotonic functions. This sequence of learning difficulty is further validated by the works of Byun in 1996 and Carroll in 1963.

Moreover, a proclivity towards linear extrapolation, even in cases of nonlinear functions, has been documented by researchers like DeLosh, Busemeyer, and McDaniel in 1997. This behavioral tendency has been interpreted through two primary classes of computational models: similarity-based models and rule-based models. The Associative Learning Model (ALM), a representative of the former, suggests that individuals memorize specific input-output examples and base predictions for new inputs on the likeness to these recorded instances. These models, however, grapple with the challenge of explaining accurate extrapolation, a crucial aspect of function learning.

On the flip side, rule-based models, as proposed by Brehmer in 1974 and Carroll in 1963, posit that individuals abstract a functional rule encapsulating consistency across examples. These models exhibit a superior capability for accurate extrapolation but often fall short in predicting human performance in function learning tasks, indicating the necessity for a more comprehensive model.

In response to this need, hybrid models like the Extrapolation-Association Model (EXAM) designed by DeLosh et al., fuse elements of similarity-based learning and basic rule-based extrapolation. This model archives input-output examples like the ALM but applies a linear rule for extrapolation beyond these instances. Comparative studies, such as the one conducted by McDaniel and Busemeyer in 2005, hint that this hybrid approach best portrays overall human function learning performance.

The study by DeLosh et al. manipulated a critical factor, the 'density' of training, referring to the number of unique stimulus-response pairs presented during training. Despite this manipulation, the density did not significantly impact the participants' performance during training or the subsequent transfer phase. This outcome indicates that the ability to extrapolate and abstract functions in human learners could be less influenced by the volume of training data and more by the underlying cognitive processes involved in function learning.

More recent models, like the Population of Linear Experts (POLE) proposed by Kalish, Lewandowsky, and Kruschke in 2004, suggest that individuals break down complex functions into local linear experts. This mixture-of-experts approach accounts for phenomena such as knowledge partitioning but struggles in predicting extrapolation without fitting to the test data. Despite these newer models, hybrid strategies like EXAM continue to provide the most complete account of human function learning and generalization, particularly in relation to extrapolation.

The study by McDaniel et al. (2009) directly contrasts the POLE and EXAM models. They trained participants on piecewise linear functions, with one segment sparsely trained. This setup put the models to test on extrapolating from sparse data. Additionally, they based model comparisons on a priori predictions rather than post-hoc fits. Parameters were fit to just the training data, and models independently predicted unseen transfer stimuli. This approach offers more diagnostic tests than fitting the models simultaneously to all data. The crucial findings reveal that while POLE provides a better quantitative fit to the training data, EXAM more accurately predicts human performance on essential extrapolation trials. For instance, POLE tends to erroneously predict that extrapolation will follow the slope of the densely trained segment, rather than the slope of the sparsely trained region as exhibited by humans. In contrast, EXAM, with its exemplar representations and linear extrapolation rule, better represents the human tendency to extrapolate along the slope defined by the sparse training data.

>>>>

In project 1, I applied model-based techniques to quantify and control for the similarity between training and testing experience, which in turn enabled us to account for the difference between varied and constant training via an extended version of a similarity based generalization model. In project 2, we will go a step further, implementing a full process model capable of both 1) producing novel responses and 2) modeling behavior in both the learning and testing stages of the experiment. Project 2 also places a greater emphasis on extrapolation performance following training. Although varied training has often been purported to be particularly beneficial for generalization or transfer, few experiments have compared varied and constant training in contexts with unambiguous extrapolation testing.

## Function Learning and Extrapolation

The study of human function learning investigates how people learn relationships between continuous input and output values.  Function learning is studied both in tasks where individuals are exposed to a sequence of input/output pairs [@deloshExtrapolationSineQua1997; @mcdanielEffectsSpacedMassed2013], or situations where observers are presented with a an incomplete scatterplot or line graph and make predictions about regions of the plot that don't contain data [@ciccioneCanHumansPerform2021a; @courrieuQuickApproximationBivariate2012; @saidExtrapolationAccuracyUnderestimates2021;@schulzCommunicatingCompositionalPatterns2020].

@carrollFunctionalLearningLearning1963 conducted the earliest work on function learning. Input stimuli and output responses were both lines of varying length. The correct output response was related to the length of the input line by a linear, quadratic, or random function.  Participants in the linear and quadratic performed above chance levels during extrapolation testing, with those in the linear condition performing the best overall. Carroll argued that these results were best explained by a ruled based model wherein learners form an abstract representation of the underlying function. Subsequent work by @brehmerHypothesesRelationsScaled1974,testing a wider array of functional forms, provided further evidence for superior extrapolation in tasks with linear functions. Brehmer argued that individuals start out with an assumption of a linear function, but  given sufficient error will progressively test alternative hypothesis with polynomials of greater degree. @kohFunctionLearningInduction1991 employed a visuomotor function learning task, wherein participants were trained on examples from an unknown function relating the length of an input line to the duration of a response (time between keystrokes). In this domain, participants performed best when the relation between line length and response duration was determined by a power, as opposed to linear function. Koh & Meyer developed the log-polynomial adaptive-regression model to account for their results.

The first significant challenge to the rule-based accounts of function learning was put forth by @deloshExtrapolationSineQua1997 .  In their task, participants learned to associate stimulus magnitudes with response magnitudes that were related via either linear, exponential, or quadratic function. Participants approached ceiling performance by the end of training in each function condition, and were able to correctly respond in interpolation testing trials. All three conditions demonstrated some capacity for extrapolation, however participants in the linear condition tended to underestimate the true function, while exponential and quadratic participants reliably overestimated the true function on extrapolation trials. Extrapolation and interpolation performance are depicted in @fig-delosh-extrap.

The authors evaluated both of the rule-based models introduced in earlier research (with some modifications enabling trial-by-trial learning). The polynomial hypothesis testing model [@carrollFunctionalLearningLearning1963; @brehmerHypothesesRelationsScaled1974] tended to mimic the true function closely in extrapolation, and thus offered a poor account of the human data.  The log-polynomial adaptive regression model [@kohFunctionLearningInduction1991] was able to mimic some of the  systematic deviations produced by human subjects, but also predicted overestimation in cases where underestimation occurred.

The authors also introduced two new function-learning models. The Associative Learning Model (ALM) and the extrapolation-association model (EXAM). ALM is a two layer connectionist model adapted from the ALCOVE model in the category learning literature [@kruschkeALCOVEExemplarbasedConnectionist1992]. ALM belongs to the general class of radial-basis function neural networks, and can be considered a similarity-based model in the sense that the nodes in the input layer of the network are activated as a function of distance. The EXAM model retains the same similarity based activation and associative learning mechanisms as ALM, while being augmented with a linear rule response mechanism. When presented with novel  stimuli, EXAM will retrieve the most similar input-output examples encountered during training, and from those examples compute a local slope. ALM was able to provide a good account of participant training and interpolation data in all three function conditions, however it was unable to extrapolate. EXAM, on the other hand, was able to reproduce both the extrapolation underestimation, as well as the quadratic and exponential overestimation patterns exhibited by the human participants. Subsequent research identified some limitations in EXAM's ability to account for cases where human participants learn and extrapolate sinusoidal function @bottNonmonotonicExtrapolationFunction2004 or to scenarios where different functions apply to different regions of the input space @kalishPopulationLinearExperts2004, though EXAM has been shown to provide a good account of human learning and extrapolation in tasks with bi-linear, V shaped input spaces @mcdanielPredictingTransferPerformance2009.

```{r}
#| label: fig-delosh-extrap
#| fig-cap: Generalization reproduced patterns from DeLosh et al. (1997) Figure 3. Stimulii that fall within the dashed lines are interpolations of the training examples. 
#| code-fold: true
#| fig-width: 10
pacman::p_load(dplyr,ggplot2,patchwork) 
source(here::here("Functions","deLosh_data.R"))
source(here::here("Functions","Display_Functions.R"))

extrapLines <- list(geom_vline(xintercept=30,color="black",alpha=.4,linetype="dashed"),
                    geom_vline(xintercept=70,color="black",alpha=.4,linetype="dashed"))

linear_function <- function(x) 2.2 * x + 30
exponential_function <- function(x) 200 * (1 - exp(-x/25))
quadratic_function <- function(x) 210 - (x - 50)^2 / 12

linear_plot <- ggplot(deLosh_data$human_data_linear, aes(x, y)) +
    geom_point(shape=1) + stat_function(fun = linear_function, color = "black") +
  labs(y="Response Magnitude", title="Linear Function",x="") + extrapLines

exponential_plot <- ggplot(deLosh_data$human_data_exp, aes(x, y)) +
  geom_point(aes(shape = "Observed", color = "Observed"),shape=1) + 
  stat_function(aes(color = "True Function"),fun = exponential_function, geom="line")+
  labs(x="Stimulus Magnitude", title="Exponential Function",y="")  +
  extrapLines +
  scale_shape_manual(values = c(1)) +
  scale_color_manual(values = c("Observed" = "black", "True Function" = "black")) +
  theme(legend.title = element_blank(), legend.position="top") +
  guides(color = guide_legend(override.aes = list(shape = c(1, NA), 
                                                  linetype = c(0, 1))))

quadratic_plot <- ggplot(deLosh_data$human_data_quad, aes(x = x, y = y)) +
  geom_point( shape = 1) +
  stat_function( fun = quadratic_function, geom = "line") +
  labs(title="Quadratic Function",x="",y="") + extrapLines

linear_plot + exponential_plot + quadratic_plot
```
